Solutions to the Assignment - 1 : CS6230 - \\ Optimization Methods in Machine Learning

### 2018-09-30 Assignment-1/Solns-Assgn-1.tex
Port remaining solutions to LaTeX

\documentclass{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}

@@

@@

@@

Now note that the eigenvalues of \(\nabla^{2} f(\mathbf{x})\) are \(\Lambda = \lbrace\frac{1}{x^{2}_{1}}, \frac{1}{x^{2}_{2}}, \ldots, \frac{1}{x^{2}_{n}}\rbrace\). Since the domain of \(f\) is \(\mathbb{R}^{n}_{++}\), we can most certainly assure that \(\nexists \lambda \in \Lambda\), such that \(\lambda \leq 0\). Hence it can be said that \(f\) is \(k\)-strongly convex with \(k = \min \Lambda\).
\end{flushleft}

>>>>

\documentclass{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{hyperref}

\newcommand{\tr}{\mathrm{tr}}
\newcommand{\deter}{\mathrm{det}}

@@

\section*{Question 3}
\begin{flushleft}
Consider an affine transformation of \(x\) given by matrices \(A \in \mathbb{R}^{k x n}\) and \(b \in \mathbb{R}^{k}\). We know that the domain of \(g(x)\) is \(\mathbb{R}^{n}\), which is convex. Note that \(Ax\) is a matrix-vector product given by:
\begin{equation}
Ax = \begin{bmatrix} a_{1}^{T}x \\ a_{2}^{T}x \\ \vdots \\ a_{k}^{T}x \end{bmatrix}
\end{equation}
Hence, \(y = Ax + b\) can be given by \(y_{i} = a_{i}^{T}x + b_{i}\quad i \in \{1, 2, \ldots, k\}\). Using a result in Boyd and Vanderberghe (Pg. 74), the canonical log-sum-exp function - \(f(x)\) is convex in \(x\).
Now the composition of a convex and a convex function is convex: therefore \(g(x) = f(y(x)) = f(Ax + b)\) is convex.
\(\newline\)

An alternative solution would be to compute the Hessian of the function. First we have to compute the gradient. Denote \(\exp(a_{i}^{T}x + b_{i}) = E_{i}\):
\begin{equation}
\frac{\partial g}{\partial x_{j}} = \frac{\displaystyle \sum_{i=1}^{n} \exp(a_{i}^{T}x + b_{i})a_{ij}}{\displaystyle \sum_{i=1}^{k} \exp(a_{i}^{T}x + b_{i})} = \frac{\displaystyle \sum_{i=1}^{n} E_{i}a_{ij}}{\displaystyle \sum_{i=1}^{n} E_{i}}
\end{equation}

For a critical point:
\begin{equation}
\label{3-grad}
\frac{\partial g}{\partial x_{q}} = \frac{\displaystyle \sum_{i=1}^{n} E_{i}a_{iq}}{\displaystyle \sum_{i=1}^{n} E_{i}} = 0
\end{equation}

Now compute the \(p;q^{th}\) element of the Hessian (ran out of symbols).
\begin{equation}
\frac{\partial^{2} g}{\partial x_{p} \partial x_{q}} = \underbrace{\frac{\displaystyle \sum_{i=1}^{k}E_{i}a_{iq}\sum_{i=1}^{k}E_{i}a_{ip}}{\displaystyle \sum_{i=1}^{k} (E_{i})^{2}}}_{= 0 \text{ due to Eqn \ref{3-grad}}} + \frac{\displaystyle \sum_{i=1}^{k}E_{i}a_{iq}a_{ip}}{\displaystyle \sum_{i=1}^{k} E_{i}}
\end{equation}

Now since the other term is only left, we have to show \(z^{T}Hz = \sum_{p,q=1}^{n}z_{p}z_{q}H_{pq} \geq 0\) to show that the function is convex.
\begin{equation}
\frac{n^2}{\displaystyle \sum_{i=1}^{k} E_{i}}\sum_{p=1}^{n}\sum_{q=1}^{n}\sum_{i=1}^{k}z_{p}z_{q}E_{i}a_{ip}a_{iq} = \frac{n^2}{\displaystyle \sum_{i=1}^{k} E_{i}}\sum_{i=1}^{k}E_{i}\sum_{p=1}^{n}z_{p}a_{ip}\sum_{q=1}^{n}z_{q}a_{iq} = \frac{n^2}{\displaystyle \sum_{i=1}^{k} E_{i}}\sum_{i=1}^{k}E_{i}\left(\sum_{p=1}^{n}z_{p}a_{ip}\right)^{2} \geq 0
\end{equation}

The last inequality comes due to the fact that \(E_{i} > 0 \forall i\), and \(x^{2} \geq 0 \forall x\).
\end{flushleft}

@@

\section*{Question 5}
\subsection*{Question 5.a.}
\begin{flushleft}
We adopt a similar proof technique as used in Boyd and Vanderberghe (Pg. 74).
\(\newline\)

Consider \(X(t) = Z + tV\) where \(Z \succ 0\) and \(V \in S^{n}\) (symmetric matrices). Let \(t\) be the set of all values such that \(X(t)\) is PD.

Now since \(X(t)\) is PD, \(X(t)^{-1}\) exists, and it given by:
\begin{equation}
X(t)^{-1} = [Z + tV]^{-1} = (Z[I + tZ^{-1}V])^{-1} = [I + tZ^{-1}V]^{-1}Z^{-1}
\end{equation}

A result in Matrix perturbations states that (analogous to a negative binomial expansion):
\begin{multline}
(I + tZ^{-1}V)^{-1} = I + tZ^{-1}V + t^{2}Z^{-1}VZ^{-1}V + \ldots\\\Rightarrow (I + tZ^{-1}V)^{-1}Z^{-1} = Z^{-1} + tZ^{-1}VZ^{-1} + t^{2}Z^{-1}VZ^{-1}VZ^{-1} + \ldots
\end{multline}

Let \(g(t)\) be given by:
\begin{equation}
g(t) = \tr(X(t)^{-1}) = \tr(Z^{-1}) + t\tr(Z^{-1}VZ^{-1}) + t^{2}\tr(Z^{-1}VZ^{-1}VZ^{-1}) + \ldots
\end{equation}

Now, compute the derivative of \(g(t)\) w.r.t. \(t\):
\begin{gather}
\frac{\partial g(t)}{\partial t} = \tr(Z^{-1}VZ^{-1}) + 2t\tr(Z^{-1}VZ^{-1}VZ^{-1}) + \ldots \\
\Rightarrow \frac{\partial^{2} g(t)}{\partial t^{2}} = 2\tr(Z^{-1}VZ^{-1}VZ^{-1}) + \ldots \\
\Rightarrow \frac{\partial^{2} g(t)}{\partial t^{2}} \vert_{t=0} = 2\tr(Z^{-1}VZ^{-1}VZ^{-1})
\end{gather}

Denote \(W = VZ^{-1}\). Hence:
\begin{equation}
\frac{\partial^{2} g(t)}{\partial t^{2}} \vert_{t=0} = 2\tr(Z^{-1}VZ^{-1}VZ^{-1}) = 2\tr(W^{T}Z^{-1}W)
\end{equation}

If \(Z\) is PD, then \(Z^{-1}\) is PD, and \(Z^{-1}V\) and \(VZ^{-1}\) are PD. Hence: \(\frac{\partial^{2} g(t)}{\partial t^{2}} \vert_{t=0} = 2\tr(W^{T}Z^{-1}W) \geq 0\), since the product of PD matrices is PSD. This means that \(g(0) = \tr(Z)\) is convex and \(Z \in S_{n}^{++}\).
\end{flushleft}

\subsection*{Question 5.b.}
\begin{flushleft}
Using a similar approach to 5.a, consider \(Y(t) = Z + tV\) where \(Z \succ 0\) and \(V \in S^{n}\).
\begin{multline}
Y(t) = Z + tV = Z^{0.5}[I + tZ^{-0.5}VZ^{-0.5}]Z^{0.5} \Rightarrow \deter(Y(t)) = \deter(Z^{0.5})\deter([I + tZ^{-0.5}VZ^{-0.5}])\deter(Z^{0.5})\\= \deter(Z)\deter([I + tZ^{-0.5}VZ^{-0.5}]).
\end{multline}

Consider the eigenvalues of \(Z^{-0.5}VZ^{-0.5}\) to be \(\{\lambda_{1}, \lambda_{2}, \ldots, \lambda_{n}\}\). Hence the eigenvalues of \(I + tZ^{-0.5}VZ^{-0.5}\) are \(\{1 + t\lambda_{1}, 1 + t\lambda_{2}, \ldots, 1 + t\lambda_{n}\}\).
\(\newline\)

Let \(h(t) = \deter(Z)^{\frac{1}{n}}\deter([I + tZ^{-0.5}VZ^{-0.5}])^{\frac{1}{n}}\):
\begin{equation}
h(t) = \deter(Z)^{\frac{1}{n}}\deter([I + tZ^{-0.5}VZ^{-0.5}])^{\frac{1}{n}} = \deter(Z)^{\frac{1}{n}}\left(\prod_{i=1}^{n}(1 + t\lambda_{i})\right)^{\frac{1}{n}}
\end{equation}

The second term involving the product is a geometric mean. Citing a result in Boyd and Vanderberghe about the geometric mean being concave for positive numbers, we show that \(h(t)\) is concave \(\forall t \in T\). The proof is complete by considering \(t = 0\).
\end{flushleft}

\section*{Question 7}
\begin{flushleft}
Consider the derivative of the function \(f(z) = |z|\) to be
\begin{equation}
f'(z) = \begin{cases} 1 & \text{ if} z \geq 0 \\ -1 & \text{ otherwise} \end{cases}
\end{equation}

Therefore, the Hessian can be written as:
\begin{equation}
\begin{bmatrix}
2a & \pm 1 \\ \pm 1 & 2a
\end{bmatrix}
\end{equation}

We need the eigenvalues to be \(> 0\), hence by finding the characteristic polynomial and extracting the roots we get \(a > \pm 0.5\). We take the intersection of the two sets, to finally get \(a > 0.5\).
\end{flushleft}

@@

Now note that the eigenvalues of \(\nabla^{2} f(\mathbf{x})\) are \(\Lambda = \lbrace\frac{1}{x^{2}_{1}}, \frac{1}{x^{2}_{2}}, \ldots, \frac{1}{x^{2}_{n}}\rbrace\). Since the domain of \(f\) is \(\mathbb{R}^{n}_{++}\), we can most certainly assure that \(\nexists \lambda \in \Lambda\), such that \(\lambda \leq 0\). Hence it can be said that \(f\) is strictly convex.
\end{flushleft}
